The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-12 rational.tex

In a recent paper~\cite{infeffects}, we argued that, because they include
irreversible physical primitives, conventional abstract models of computation
have inadvertently included some \emph{implicit} computational effects which
we called \emph{information effects}. We then developed a pure reversible
model of computation that is obtained from the type isomorphisms and
categorical structures that underlie models of linear logic and quantum
computing and that treats information as a linear resource that can neither
be erased nor duplicated. In this paper, we show that our pure reversible
model unveils deeper and more elegant symmetries of computation than have
previously been reported. In particular, we expose two notions of duality of
computation: an additive duality and a multiplicative duality that give rise
to negative types and fractional types respectively. Although these types
have previously appeared in the literature (see Sec.~\ref{sec:related}), they
have typically appeared in the context of conventional languages with
information effects, which obscured their appeal and murked their properties.

@@

\paragraph*{Contributions and Outline.} 
To summarize, in a computational framework that guarantees that information
is preserved, negative and fractional types provide fascinating mechanisms in
which computations can be sliced and diced, decomposed and recomposed, run
forwards and backwards, in arbitrary ways. The remainder of the paper
formalizes these informal observations. Specifically our main contributions
are:
\begin{itemize}
\item We extend {{langRev}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} (reviewed in Sec.~\ref{sec:pi}) with a notion
  of negative types, that satisfies the isomorphism {{a + (-a) <-> 0}}. The
  semantics of this extension is expressed by having a \emph{dual} evaluator
  that reverses the flow of execution for negative
  values. (Sec.~\ref{sec:neg})
\item We independently extend {{langRev}} with a notion of fractional types,
  that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
  extension is expressed by introducing logic variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types. (Sec.~\ref{sec:frac})
\item We combine the above two extensions into a language whose type system
  allows any rational number to be used as a type. Moreover the types satisfy
  the same familiar and intuitive isomorphisms that are satisfied in the
  mathematical field of rational numbers. (Sec.~\ref{sec:rat})
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every
  programmer. (Sec.~\ref{sec:prog})
\item We relate our notions of negative and fractional types to previous work
  on continuations. Briefly, we argue that conventional continuations
  conflate negative and fractional components. This observation allows us to
  relate two apparently unrelated lines of work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relating continuations to
  negative types and the second~\cite{Bernardi:2010:CSL:1749618.1749689}
  relating continuations to the fractional types of the Lambek-Grishin
  calculus. (Sec.~\ref{sec:related})
\end{itemize}

@@

\item Sum types may similarly be represented by one wire or using using
  parallel wires with a {{+}} operator between them. When tracing the
  execution of two additive wires, a value can reside on only one of the two
  wires.
\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
\includegraphics{diagrams/thesis/sum-one-wire.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
}
\end{center}
\end{multicols}
\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
\includegraphics{diagrams/thesis/sum-two-wires.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires-right-value.pdf}
}
\end{center}
\end{multicols}

@@

This section introduces the syntax and graphical languages for
negative types and fractionals.The general outline is as follows:
\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphims traditionally called {{eta}} and {{eps}}. 
\end{itemize}
After presenting the extension at the syntactic level, we discuss the
categorical semantics informally via the graphical language. The operational
semantics is presented in Sec.~\ref{sec:rat}.

@@

For convenience we sometime use the notations {{b1 - b2}} and {{b1/b2}} to
indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}} respectively.  The
types of the new constructs are: 
\vspace{-15pt}
\begin{multicols}{2}  
%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+
% eta* &: 1 <-> (1/b) * b :& eps*

@@

For a monoidal category to be compact closed with a dual objects
presented as $b^*$, the maps~{{eta}} and~{{eps}} must satisfy a
coherence condition that is usually visualized as follows:
\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}
In the case of negatives, the condition amounts to checking that
reversing direction twice is a no-op. In the case of fractionals, the
condition amounts to checking that creating values {{alpha}} and
{{1/alpha}} and immediately unifying them is also a no-op. Both checks
are straightforward and are essentially the constructions in the
introduction.

@@

\paragraph*{Trace.}
Every compact closed categoy admits a trace. For the additive case, we get
the following definition.  Given {{f : b1+b2 <-> b1+b3 }}, 
define {{trace+ f : b2 <-> b3}} as:

@@

\noindent We have omitted some the commutativity and associativity shuffling to
communicate the main idea. We are given a value of type {{b2}} which we embed
into {{0+b2}} and then {{(-b1+b1)+b2)}}. This can be reassociated into 
{{-b1+(b1+b2)}}. The component {{b1+b2}} which until now is just an appropriately 
tagged value of type {{b2}} can be given to~{{f}} which transforms it to
something of type {{b1+b3}}. If the result is in the {{b3}}-summand, it is produced
as the answer; otherwise the result is in the {{b1}}-summand; {{eps+}} is used to
make it flow backwards to be fed to the {{eta+}} located at the beginning of the 
sequence. Iteration continues until a {{b3}} is produced.

@@

The multiplicative construction is similar. This is a distinguishing
difference from *-autonomous categories that are models for Linear
Logic, where the dual does not preserve the tensor but maps to a dual
tensor \cite{curien2009interactive}.

@@

The {{inv}} construction is similar.  We are able to do this because
in {{langRevEE}} every morphism is an isomorphism and hence has an
inverse morphism, whereas in compact closed categories with no
underlying dagger structure we can only construct an op-functor.

@@

\paragraph*{Lifting negation out of {{*}}.}
The isomorphisms below state that the direction is \emph{relative}. If {{b1}}
and {{b2}} are flowing opposite to each other then it doesn't matter which is
direction is forwards and which is backwards. More interestingly as {{b1}} is
moving backwards, it can ``see the past'' of {{b2}} which is equivalent to
both particles moving backwards.

@@

To build these isomorphisms, we first build an intermeidate
construction which we call {{eps_{fst} : (-b1)*b2 + b1*b2 <-> 0}}. 
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/eps_fst.pdf}
}
\end{center}

@@

In this section we develop an operational semantics for {{langRev}} extended
with negative and fractional types, {{langRevEE}}. The operational semantics
takes the abstract categorical diagrams which were previously known but gives
them a computational interpretation based on reverse execution for negative
types and unification for fractional types. Even though this computational
interpretation appears straightforward in retrospect, it constitutes a
breakthrough because it breaks the preconceived idea that there is only one
notion of duality and hence that both negative and fractional types should be
implemented using the same underlying mechanism. In some sense, the folkore
literature on monoidal categories describing the evaluations for additive and
multiplicative fragments are ``particle-style'' and ``wave-style'' can be
seen to suggest either a unified or separate implementations. But, since de
Morgan, we have been predisposed to think of only one notion of duality which
persisted to even linear logic. (See Sec.~\ref{sec:related} for further
discussion of duality.)

@@

\begin{enumerate}
\item We rewrite the semantics in Sec~\ref{sec:pi} using unification of the
  values instead of direct pattern matching. This gives us the neccessary
  infrastructure for fractional types. We assume the reader is familiar with
  the idea of unification as realized using logic variables, substitutions,
  and reification as presented in any standard text on logic programming.
\item We write a reverse interpreter by reversing the direction of the
  {{|-->}} reductions. This gives us the neccessary infrastructure for
  negative types.
\end{enumerate}

@@

Using the same the entire semantics can be extended to thread the
substitution as shown below:

@@

  \begin{enumerate}
  \item The rules for {{eps+}} essentially transfer control from the
    forward evaluator (whose states are tagged by $\triangleright$) to
    the backward evaluator (whose states are tagged by
    $\triangleleft$). In other words, after an {{eps+}} the direction
    of the world is reversed. The pattern matching done by the
    unification ensures that a value on the {{right}} wire is tagged
    to be negative and transfered to the {{left}} wire, and vice versa.
%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eps+, v, C, s>_{fwd} |--> <eps+, left (-v'), C, s[v <><> right v']>_{bck}
% <eps+, v, C, s>_{fwd} |--> <eps+, right v', C, s[v <><> left (-v')]>_{bck}
   %
    Note that there is no evaluation rule for {{eta+}} in the forward
    evaluator. This corresponds to the fact that there is no value of
    type {{0}} and hence the forward evaluator can never execute an
    {{eta+}}.

    \item The rules for {{eta+}} are added to the backward evaluator. A
      program executing backward starts executing forwards after the
      execution of the {{eta+}}. Dual to the previous case, there is
      no rule for {{eps+}} in the backward evaluator since the output
      type of {{eps+}} is {{0}}.
%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eta+, v, C, s>_{bck} |--> <eta+, left (-v'), C, s[v <><> right v']>_{fwd}
% <eta+, v, C, s>_{bck} |--> <eta+, right v', C, s[v <><> left (-v')]>_{fwd}

@@

This construction can be used to solve a large class of constraint
satisfaction problems. We illustrate the main ideas with the implementation
of a SAT-solver. Although the main idea is the use of {{trace*}}, the
construction is involved so we proceed in small steps, reviewing some of the
necessary constructions presented in our earlier paper~\cite{infeffects}.

@@

The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
negates its second argument if the first argument is {{true}}. This
gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
equivalent construction that is useful is {{else_{not} }} where we
negate the second argument only is the first is {{false}}.

@@

Although cloning is generally not allowed in reversible languages, it is
possible at the cost of having additional constant inputs. For example,
consider the gate {{else_{not} }}. Generally, the gate gates maps
{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
on the cases in which the second input is {{true}}, we get that the gate maps
{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
i.e., the gate clones the first input. A circuit of {{n}} parallel
{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
{{true}} inputs in the process.  Let us call this construction
{{clone^n_{bool} }}.

@@

With a little work, we can modify this circuit to only annihilate values that
fail to satisfy the constraints reprsented by a SAT-instance {{f}}. In more
detail, an instance of SAT is a function/circuit~{{f}} that given some
boolean inputs returns {{true}} or {{false}} which we interpret as whether
the inputs satisfy the constraints imposed by the structure of {{f}}. Because
we are in a reversible world, our instance of SAT must be expressed as an
isomorphism: this is easily achieved as shown in Sec.~\ref{sub:f}
below. Assuming that {{f}} is expressed as an isomorphism, we have enough
information to reconstruct the input from the output. This can be done by
using the adjoint of {{f}}. At this point we have, the top half of the
construction below:

@@

Any boolean expression {{f : bool^n -> bool}} can be compiled into the
isomorphism {{iso : bool^h*bool^n<->bool^g*bool}} where the extra bits
{{bool^h}} and {{bool^g}} are considered as heap and garbage. Constructing
such an isomorphism has been detailed
before~\cite{Toffoli:1980,infeffects}. The important relation to note is that
applying {{iso}} to some special heap values and an input \textit{bs}
produces some bits that can be ignored and the same output that {{f}} would
have produced on {{bs}}. We can ensure that the heap has the appropriate
initial values by checking the heap and negating a second control wire, if
the values do not match, i.e., the dotted part in the diagram below.

@@

Let us call the above construction which maps inputs, heap, and control wires
to input, heap, and control wires as {{sat_f}}. The SAT-solver is is
completed by tracing the {{sat_f}} and cloning the inputs using
{{clone^n_{bool} }}.

@@

When the solver is fed inputs initialized to {{true}}, the {{clone}} will
clone any inputs to {{sat_f}} that satisfy the constraints imposed by {{f}}
and the constraints on the {{heap}}. In the case of unique-SAT the solver
will produce exactly 0 or 1 solutions. In the case of general SAT, the solver
will produce solutions as determined by the semantics of the top-level
interaction as discussed in Sec. \ref{sec:rat}.

@@

The idea of ``negative types'' has appeared many times in the literature and
has often been related to some form of continuations. Fractional types are
less common but have also appeared in relation to parsing natural
languages. Although each of these previous occurrences of negative and
fractional types is somewhat related to our work, our results are
subtantially different. To clarify this point, we start by reviewing the
salient point of the major pieces of related work and conclude this section
with a summary contrasting our approach to previous work.

@@

\paragraph*{The Duality of Computation.}
The duality between call-by-name and call-by-value was further investigated
by Selinger using control
categories~\cite{Selinger:2001:CCD:966910.966911}. Curien and
Herbelin~\cite{Curien:2000} also introduce a calculus that exhibits
symmetries between values and continuations and between call-by-value and
call-by-name. The calculus includes the type $A-B$ which is the dual of
implication, i.e., a value of type $A-B$ is a context expecting a function of
type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation of type
$B$. This is to be contrasted with our interpretation of a value of that type
as \emph{either} a value of $A$ or a demand for a value of type $B$. This
calculus was further analyzed and extended by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives no
interpretation to the subtraction connective and like the original symmetric
calculus of Filinksi, introduces a duality that relates sums to products and
vice-versa.

@@

We have extended the language {{langRev}} that expressed computation
in the commutative semiring of whole numbers to {{langRevEE}} that
expresses computation in the fields of rationals.  Every algebraic
identity that holds for the rational numbers corresponds to a type
isomorphism with a computational interpretation in our model. We have
examined the two function spaces that arise in this model and
developed non-trivial constructions such as a SAT-solver that relies
on a multiplicative trace.

In another sense however, this paper is about the nature of duality in
computation. 
% The concept of duality is deep and significant and in many cases ---
% probably most famously in the divide between classical and intuitionistic
% mathematics --- contain non-trivial assumptions about our worldview.  The
% problem of duality in computation and logic, has many facets of which
% `continuation' and `higher order functions' are only one.
The concept of duality is deep and significant: we have opened the door for
us to consider, not one but two notions of duality. Surprisingly this makes
things significantly simpler. In particular, instead of conflating
considering pairs as dual to sums, the tradition in mathematics has long been
to consider fields with two notions of duality: one for sums and one for
pairs. This double notion of duality has a crisp semantics, clear
computational interpretation, and an information theoretic basis.

Our work has barely scratched the surface of an area of computing which has
been explored in depth before but without the combined reversible
information-preserving framework and the two notions of duality. The new
insights point to further new areas of investigation of which we mention
the three most significant ones (in our opinion).

@@

We have confirmed that we can extended {{langRevEE}} with the datatype
declaration for binary trees and build a witness for this isomorphism that
works as expected. To understand the issue in more detail, consider the
following algebraically valid proof of the isomorphism in question:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
The question is why such an algebraic manipulation makes sense type
theoretically, even though the intermediate step asking for an isomorphism
between $x^6$ and $1$ has no computational context. In the setting of
{{langRevEE}}, this isomorphism can be constructed but it diverges on all
inputs (in both ways). This suggests that, in the field of algebraic numbers,
some algebraic manipulations are somehow 
``more constructive'' than others.
%% or dually that {{langRevEE}} with recursive types
%%somehow lacks the ability to express all computations that are otherwise
%%algebraically meaningful.

@@

\paragraph*{Quantum Computing.} 
One understand of quantum computing is that it exploits the laws of physics
to build faster machines (perhaps). Another more foundational understanding
is that it provides a computational interpretation of physics, and in
particular directly address the question of interpretation of quantum
mechanics. In a little known document, Rozas~\cite{Rozas:1987:CMO:889539}
uses continuations to implement the transactional interpretation of quantum
mechanics~\cite{transactional} which includes as its main ingredient a
fixpoint calculation between waves or particles traveling forwards and
backwards in time. Our work shed no light on whether this interpretation is
the ``right one'' but it is interesting that we can directly realize it using
the primitives of {{langRevEE}}.

>>>>

In a recent paper~\cite{infeffects}, we argued that, because they include
irreversible physical primitives, conventional abstract models of computation
have inadvertently included some \emph{implicit} computational effects which
we called \emph{information effects}. We then developed a pure reversible
model of computation that is obtained from the type isomorphisms and
categorical structures that underlie models of linear logic and quantum
computing and that treats information as a linear resource that can neither
be erased nor duplicated. In this paper, we show that our pure reversible
model unveils deeper and more elegant symmetries of computation than have
previously been reported. In particular, we expose two notions of duality of
computation: an additive duality and a multiplicative duality that give rise
to negative types and fractional types respectively. Although these types
have previously appeared in the literature (see Sec.~\ref{sec:related}), they
have typically appeared in the context of conventional languages with
information effects, which limited their appeal and obscured their properties.

@@

\paragraph*{Contributions and Outline.} 
To summarize, in a computational framework that guarantees that information
is preserved, negative and fractional types provide fascinating mechanisms in
which computations can be sliced and diced, decomposed and recomposed, run
forwards and backwards, in arbitrary ways. The remainder of the paper
formalizes these informal observations. Specifically our main contributions
are:
\begin{itemize}
\item We extend {{langRev}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} (reviewed in Sec.~\ref{sec:pi}) with a notion
  of negative types, that satisfies the isomorphism {{a + (-a) <-> 0}}. The
  semantics of this extension is expressed by having a \emph{dual} evaluator
  that reverses the flow of execution for negative
  values. (Sec.~\ref{sec:neg})
\item We independently extend {{langRev}} with a notion of fractional types,
  that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
  extension is expressed by introducing logic variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types. (Sec.~\ref{sec:frac})
\item We combine the above two extensions into a language, which we
  call {{langRevEE}}, whose type system allows any rational number to
  be used as a type. Moreover the types satisfy the same familiar and
  intuitive isomorphisms that are satisfied in the mathematical field
  of rational numbers. (Sec.~\ref{sec:rat})
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every
  programmer. (Sec.~\ref{sec:prog})
\item We relate our notions of negative and fractional types to previous work
  on continuations. Briefly, we argue that conventional continuations
  conflate negative and fractional components. This observation allows us to
  relate two apparently unrelated lines of work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relating continuations to
  negative types and the second~\cite{Bernardi:2010:CSL:1749618.1749689}
  relating continuations to the fractional types of the Lambek-Grishin
  calculus. (Sec.~\ref{sec:related})
\end{itemize}

@@

\item Sum types may similarly be represented by one wire or using
  parallel wires with a {{+}} operator between them. When tracing the
  execution of two additive wires, a value can reside on only one of the two
  wires.
\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
\includegraphics{diagrams/thesis/sum-one-wire.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
}
\end{center}
\end{multicols}
\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
\includegraphics{diagrams/thesis/sum-two-wires.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires-right-value.pdf}
}
\end{center}
\end{multicols}

@@

This section introduces the syntax and graphical languages for
negative types and fractionals.The general outline is as follows:
\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphisms traditionally called {{eta}} and {{eps}}. 
\end{itemize}
After presenting the extension at the syntactic level, we discuss the
categorical semantics informally via the graphical language. The operational
semantics is presented in Sec.~\ref{sec:rat}.

@@

For convenience, we sometimes use the notations {{b1 - b2}} and {{b1/b2}} to
indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}} respectively.  The
types of the new constructs are: 
\vspace{-15pt}
\begin{multicols}{2}  
%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+
% eta* &: 1 <-> (1/b) * b :& eps*

@@

For a monoidal category to be compact closed the maps~{{eta}}
and~{{eps}} must satisfy a coherence condition that is usually
visualized as follows:
\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}
where $b^*$ represents the dual of {{b}}.  In the case of negatives,
the condition amounts to checking that reversing direction twice is a
no-op. In the case of fractionals, the condition amounts to checking
that creating values {{alpha}} and {{1/alpha}} and immediately
unifying them is also a no-op. Both checks are straightforward and are
essentially the constructions in the introduction.

@@

\paragraph*{Trace.}
Every compact closed category admits a trace. For the additive case, we get
the following definition.  Given {{f : b1+b2 <-> b1+b3 }}, 
define {{trace+ f : b2 <-> b3}} as:

@@

\noindent We have omitted some of the commutativity and associativity
shuffling to communicate the main idea. We are given a value of type
{{b2}} which we embed into {{0+b2}} and then {{(-b1+b1)+b2)}}. This
can be re-associated into {{-b1+(b1+b2)}}. The component {{b1+b2}},
which until now is just an appropriately tagged value of type {{b2}},
is transformed to a value of type {{b1+b3}} by {{f}}. 
If the result is in the {{b3}}-summand, it is produced as
the answer; otherwise the result is in the {{b1}}-summand; {{eps+}} is
used to make it flow backwards to be fed to the {{eta+}} located at
the beginning of the sequence. Iteration continues until a {{b3}} is
produced.

@@

\noindent
The multiplicative construction is similar.

%% This is a distinguishing difference from *-autonomous categories that
%% are models for Linear Logic, where the dual does not preserve the
%% tensor but maps to a dual tensor \cite{curien2009interactive}.

@@

This construction relies on the fact that every {{langRevEE}} morphism
has an adjoint.  The {{inv}} construction is similar.

%% inverse morphism, whereas in compact closed categories with no
%% underlying dagger structure we can only construct an op-functor.

@@

\paragraph*{Lifting negation out of {{*}}.}
The isomorphisms below state that the direction is \emph{relative}. If {{b1}}
and {{b2}} are flowing opposite to each other then it doesn't matter which
direction is forwards and which is backwards. More interestingly as {{b1}} is
moving backwards, it can ``see the past'' of {{b2}} which is equivalent to
both particles moving backwards.

@@

To build these isomorphisms, we first build an intermediate
construction which we call {{eps_{fst} : (-b1)*b2 + b1*b2 <-> 0}}. 
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/eps_fst.pdf}
}
\end{center}

@@

In this section we develop an operational semantics for {{langRev}} extended
with negative and fractional types, {{langRevEE}}. The operational semantics
takes the abstract categorical diagrams which were previously known but gives
them a computational interpretation based on reverse execution for negative
types and unification for fractional types. Even though this computational
interpretation appears straightforward in retrospect, it constitutes a
breakthrough because it breaks the preconceived idea that there is only one
notion of duality and hence that both negative and fractional types should be
implemented using the same underlying mechanism. In some sense, the folklore
literature on monoidal categories describing the evaluations for additive and
multiplicative fragments are ``particle-style'' and ``wave-style'' can be
seen to suggest either a unified or separate implementations. But, since de
Morgan, we have been predisposed to think of only one notion of duality which
persisted to even linear logic. (See Sec.~\ref{sec:related} for further
discussion of duality.)

@@

\begin{enumerate}
\item We rewrite the semantics in Sec~\ref{sec:pi} using unification of the
  values instead of direct pattern matching. This gives us the necessary
  infrastructure for fractional types. We assume the reader is familiar with
  the idea of unification as realized using logic variables, substitutions,
  and reification as presented in any standard text on logic programming.
\item We write a reverse interpreter by reversing the direction of the
  {{|-->}} reductions. This gives us the necessary infrastructure for
  negative types.
\end{enumerate}

@@

Using the same idea, the entire semantics can be extended to thread
the substitution as shown below:

@@

  \begin{enumerate}
  \item The rules for {{eps+}} essentially transfer control from the
    forward evaluator (whose states are tagged by $\triangleright$) to
    the backward evaluator (whose states are tagged by
    $\triangleleft$). In other words, after an {{eps+}} the direction
    of the world is reversed. The pattern matching done by the
    unification ensures that a value on the {{right}} wire is tagged
    to be negative and transferred to the {{left}} wire, and vice versa.
%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eps+, v, C, s>_{fwd} |--> <eps+, left (-v'), C, s[v <><> right v']>_{bck}
% <eps+, v, C, s>_{fwd} |--> <eps+, right v', C, s[v <><> left (-v')]>_{bck}
   %
    Note that there is no evaluation rule for {{eta+}} in the forward
    evaluator. This corresponds to the fact that there is no value of
    type {{0}} and hence the forward evaluator can never execute an
    {{eta+}}.

    \item The rules for {{eta+}} are added to the backward evaluator. A
      program executing backwards starts executing forwards after the
      execution of the {{eta+}}. Dual to the previous case, there is
      no rule for {{eps+}} in the backward evaluator since the output
      type of {{eps+}} is {{0}}.
%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eta+, v, C, s>_{bck} |--> <eta+, left (-v'), C, s[v <><> right v']>_{fwd}
% <eta+, v, C, s>_{bck} |--> <eta+, right v', C, s[v <><> left (-v')]>_{fwd}

@@

A large class of constraint satisfaction problems can be expressed
using {{trace*}}. We illustrate the main ideas with the implementation
of a SAT-solver.  We proceed in small steps, reviewing some of the
necessary constructions presented in our earlier
paper~\cite{infeffects}.

@@

The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
negates its second argument if the first argument is {{true}}. This
gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
equivalent construction that is useful is {{else_{not} }} where we
negate the second argument only if the first is {{false}}.

@@

Although cloning is generally not allowed in reversible languages, it is
possible at the cost of having additional constant inputs. For example,
consider the gate {{else_{not} }}. Generally, the gate maps
{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
on the cases in which the second input is {{true}}, we get that the gate maps
{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
i.e., the gate clones the first input. A circuit of {{n}} parallel
{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
{{true}} inputs in the process.  Let us call this construction
{{clone^n_{bool} }}.

@@

With a little work, we can modify this circuit to only annihilate values that
fail to satisfy the constraints represented by a SAT-instance {{f}}. In more
detail, an instance of SAT is a function/circuit~{{f}} that given some
boolean inputs returns {{true}} or {{false}} which we interpret as whether
the inputs satisfy the constraints imposed by the structure of {{f}}. Because
we are in a reversible world, our instance of SAT must be expressed as an
isomorphism: this is easily achieved as shown in Sec.~\ref{sub:f}
below. Assuming that {{f}} is expressed as an isomorphism, we have enough
information to reconstruct the input from the output. This can be done by
using the adjoint of {{f}}. At this point we have, the top half of the
construction below:

@@

Any boolean expression {{f : bool^n -> bool}} can be compiled into the
isomorphism {{iso_f : bool^h*bool^n<->bool^g*bool}} where the extra bits
{{bool^h}} and {{bool^g}} are considered as heap and garbage. Constructing
such an isomorphism has been detailed
before~\cite{Toffoli:1980,infeffects}. The important relation to note is that
applying {{iso_f}} to some special heap values and an input \textit{bs}
produces some bits that can be ignored and the same output that {{f}} would
have produced on {{bs}}. We can ensure that the heap has the appropriate
initial values by checking the heap and negating a second control wire, if
the values do not match, i.e., the dotted part in the diagram below.

@@

Let us call the above construction which maps inputs, heap, and
control wires to inputs, heap, and control wires as {{sat_f}}. The
SAT-solver is is completed by tracing the {{sat_f}} and cloning the
inputs using {{clone^n_{bool} }}.

@@

When the solver is fed inputs initialized to {{true}}, it clones only
those inputs to {{sat_f}} that satisfy {{f}} and the {{heap}}
constraints. In the case of unique-SAT the solver will produce exactly
0 or 1 solutions. In the case of general SAT, the solver will produce
solutions as determined by the semantics of the top-level interaction
(see discussion in Sec. \ref{sec:rat}).

@@

The idea of ``negative types'' has appeared many times in the literature and
has often been related to some form of continuations. Fractional types are
less common but have also appeared in relation to parsing natural
languages. Although each of these previous occurrences of negative and
fractional types is somewhat related to our work, our results are
substantially different. To clarify this point, we start by reviewing the
salient point of the major pieces of related work and conclude this section
with a summary contrasting our approach to previous work.

@@

\paragraph*{The Duality of Computation.}
The duality between call-by-name and call-by-value was further investigated
by Selinger using control
categories~\cite{Selinger:2001:CCD:966910.966911}. Curien and
Herbelin~\cite{Curien:2000} also introduce a calculus that exhibits
symmetries between values and continuations and between call-by-value and
call-by-name. The calculus includes the type $A-B$ which is the dual of
implication, i.e., a value of type $A-B$ is a context expecting a function of
type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation of type
$B$. This is to be contrasted with our interpretation of a value of that type
as \emph{either} a value of $A$ or a demand for a value of type $B$. This
calculus was further analyzed and extended by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives no
interpretation to the subtraction connective and like the original symmetric
calculus of Filinski, introduces a duality that relates sums to products and
vice-versa.

@@

We have extended the language {{langRev}} that expressed computation
in the commutative semiring of whole numbers to {{langRevEE}} that
expresses computation in the field of rationals.  Every algebraic
identity that holds for the rational numbers corresponds to a type
isomorphism with a computational interpretation in our model. We have
examined the two function spaces that arise in this model and
developed non-trivial constructions such as a SAT-solver that relies
on a multiplicative trace.

In another sense however, this paper is about the nature of duality in
computation. 
% The concept of duality is deep and significant and in many cases ---
% probably most famously in the divide between classical and intuitionistic
% mathematics --- contain non-trivial assumptions about our worldview.  The
% problem of duality in computation and logic, has many facets of which
% `continuation' and `higher order functions' are only one.
The concept of duality is deep and significant: we have opened the
door for us to consider, not one but two notions of
duality. Surprisingly this makes things substantially simpler. In
particular, instead of conflating pairs as dual to sums, the tradition
in mathematics has long been to consider fields with two notions of
duality: one for sums and one for pairs. This double notion of duality
has a crisp semantics, clear computational interpretation, and an
information theoretic basis.

Our work has barely scratched the surface of an area of computing which has
been explored in depth before but without the combined reversible
information-preserving framework and the two notions of duality. The new
insights point to further new areas of investigation, of which we mention
the three most significant ones (in our opinion).

@@

We have confirmed that we can extended {{langRevEE}} with the datatype
declaration for binary trees and build a witness for this isomorphism
that works as expected. However not every isomorphism constructed from
algebraic manipulation is computationally meaningful. To understand the
issue in more detail, consider the following algebraically valid proof
of the isomorphism in question:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
The question is why such an algebraic manipulation makes sense type
theoretically, even though the intermediate step asking for an isomorphism
between $x^6$ and $1$ has no computational context. In the setting of
{{langRevEE}}, this isomorphism can be constructed but it diverges on all
inputs (in both ways). This suggests that, in the field of algebraic numbers,
some algebraic manipulations are somehow 
``more constructive'' than others.

A related issue is that not all meaningful recursive types are
meaningful polynomials. For instance {{nat=@@x.(1+x)}} implies the
polynomial {{x=1+x}} which has no algebraic solutions without appeal
to more complex structures with limits etc.

%% or dually that {{langRevEE}} with recursive types
%%somehow lacks the ability to express all computations that are otherwise
%%algebraically meaningful.

@@

\paragraph*{Quantum Computing.} 
One understanding of quantum computing is that it exploits the laws of physics
to build faster machines (perhaps). Another more foundational understanding
is that it provides a computational interpretation of physics, and in
particular directly addresses the question of interpretation of quantum
mechanics. In a little known document, Rozas~\cite{Rozas:1987:CMO:889539}
uses continuations to implement the transactional interpretation of quantum
mechanics~\cite{transactional} which includes as its main ingredient a
fixpoint calculation between waves or particles traveling forwards and
backwards in time. Our work sheds no light on whether this interpretation is
the ``right one'' but it is interesting that we can directly realize it using
the primitives of {{langRevEE}}.
